---
title: "Week 11 Assignment"
author: "Gillian McGovern"
date: "2025-04-11"
output:
  html_document:
    df_print: paged
---

### Overview

This lab explores popular R recombination system. After doing research online, I saw that the `recommenderlab` was a recommended package to use for R.

### Load the Libraries

```{r}
library(recommenderlab)
```

### Get the Data

The below data frame is the survey data I collected:

```{r}
df <- data.frame(Id = c(1, 2, 3, 4, 5),
                 Barbie = c(5, 4, 4, 3, 4),
                 Oppenheimer = c(3, 2, 2, 3, 4),
                 TopGunMaverick = c(3, 3, 3, 5, 3),
                 AvatarTheWayOfWater = c(2, 5, 5, 4, 4),
                 Wicked = c(4, 3, 4, 3, 4),
                 GetOut = c(5, 5, 4, 4, 5))
```

I then turned the data frame into a matrix:

```{r}
movie_matrix <- as.matrix(df[, -which(names(df) == "Id")])
row.names(movie_matrix) = c("u1", "u2", "u3", "u4", "u5")
r <- as(movie_matrix, "realRatingMatrix")
r
getRatingMatrix(r)
```

### Creating the Recommender

Create the evaluation scheme using the split method:

```{r}

# eval_scheme <- evaluationScheme(r, method="cross", train=0.8,
#                       given=-1, goodRating=5)
eval_scheme <- evaluationScheme(r, method="split", train=0.9, k=3, given=-1, goodRating=5)
eval_scheme
```

Evaluate several recommender algorithms using an evaluation scheme:

```{r}
algorithms <- list("random items" = list(name="RANDOM", param=NULL),
                   "popular items" = list(name="POPULAR", param=NULL),
                   "user-based CF" = list(name="UBCF", param=list(nn=50)),
                   "item-based CF" = list(name="IBCF", param=list(k=50)),
                   "SVD approximation" = list(name="SVD", param=list(k = 3)))
results <- recommenderlab::evaluate(
  eval_scheme,
  algorithms,
  type= "topNList",
  n = 1:5)
results
```

```{r}
getConfusionMatrix(results$`random items`)[[1]]
avg(results)
```

