---
title: "Gillian McGovern - DATA 607 Assignment 2"
author: "Gillian McGovern"
date: "2025-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This assignment will explore:

* Building a Table in SQL
* Storing data in SQL database
* Transfering data from SQL database to R dataframe
* Handling missing data strategy

The table that will be used is about the ratings of six recent popular movies:

* Barbie
* Oppenheimer
* Top Gun Maverick
* Avatar The Way of Water
* Wicked
* Get Out

## Loading the Libraries

```{r}
library(RMySQL)
```


## Build Table

Connect to MySQL:

```{r}
# Connect to MySQL
con <- dbConnect(MySQL(),
                 user = "gillian.mcgovern82",
                 password = "movies607", # new password I set below
                 dbname = "gillian.mcgovern82",
                 host = "cuny607sql.mysql.database.azure.com",
                 port = 3306)
# Check if the connection is successful
if(!dbIsValid(con)) {
  stop("Connection failed")
} else {
  print("Connection successful")
}
```



```{r}
drop_table_query <- 'DROP TABLE IF EXISTS Movies'
dbGetQuery(con, drop_table_query)
create_table_query <- 'CREATE TABLE Movies(Id INTEGER PRIMARY KEY, Barbie INT, Oppenheimer INT, TopGunMaverick INT, AvatarTheWayOfWater INT, Wicked INT, GetOut INT)'
dbGetQuery(con, create_table_query)

# Check the table creation was a success
show_tables_query <- "SHOW tables"
tables_result <- dbGetQuery(con, show_tables_query)
print(tables_result)
```

## Store data in SQL database

```{r}
# Insert data into table
dbGetQuery(con,"INSERT INTO Movies VALUES(1, 5, 3, 3, 2, 4, 5)")
dbGetQuery(con,"INSERT INTO Movies VALUES(2, 4, 2, 3, 5, 3, 5)")
dbGetQuery(con,"INSERT INTO Movies VALUES(3, 4, 2, 3, 5, 4, 4)")
dbGetQuery(con,"INSERT INTO Movies VALUES(4, 3, 3, 5, 4, 3, 4)")
dbGetQuery(con,"INSERT INTO Movies VALUES(5, 3, 4, 3, 4, 4, 5)")

# Change password
dbGetQuery(con,"SET PASSWORD = 'movies607'")
```


## Transfer data from SQL database to R dataframe

```{r}
data_query <- "SELECT * FROM Movies";
df <- dbGetQuery(con, data_query)
print(df)
```

## Missing data strategy

Handling missing data most likely depends on the project, the data you're looking at and why the value(s) is missing. For this particular scenario, if someone is just filling out a questionnaire and a score is missing, I would assume that the person wasn't really paying attention to the survey itself. So I would omit the person's entire response to the survey, and therefore would remove a row from the dataframe as I do not trust that response.

For many stats in R, you can add `na.rm = TRUE` to omit the `NA` value. So for example, if I wanted to get the mean rating for Wicked but only 4 out of the 5 people rated that movie, I would do:

```{r}
partial_responses_wicked <- c(4, 3, NA, 3, 4) # Just for example, I am using a vector as opposed to pulling a column from a df
mean(x = partial_responses_wicked, na.rm = TRUE)
```

If we know the value is missing simply due to it being missed by the person taking the survey, a cool way to handle missing info would be to predict the person's rating based on other people's ratings, and the person's ratings of the other movies.

## Bonus Challenge Questions

* Are you able to use a password without having to share the password with people who are
viewing your code?

One way to do this is putting the password in a file in GitHub and reading the file. Since this is still not very secure, this would not be the ideal way to do this.

```{r}
pw_df <- read.csv(url("https://raw.githubusercontent.com/gillianmcgovern0/cuny-data-607/refs/heads/main/movies_607.csv"))
pw <- colnames(pw_df)

# Connect to MySQL
con <- dbConnect(MySQL(), 
                 user = "gillian.mcgovern82", 
                 password = pw, 
                 dbname = "gillian.mcgovern82", 
                 host = "cuny607sql.mysql.database.azure.com", 
                 port = 3306)
# Check if the connection is successful
if(!dbIsValid(con)) {
  stop("Connection failed")
} else {
  print("Connection successful")
}
```

A method I would like to try to use, but didn't have a chance, is Vault, particularly vaultr. Based what I've seen online, you could create a VAULT_AUTH_GITHUB_TOKEN envrironment variable where anyone could connect and access to vault via GitHub:

```{r}
# github_vault_client <- vaultr::vault_client(login = "github")
```

Then in Vault, you would create a secret to represent the password. That way you would just need the path to the secret and not the password itself.
