---
title: "Week 9 Assignment"
author: "Gillian McGovern"
date: "2025-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

This assignment explores how to use an API to import JSON into R, and create a data frame.

The API that I chose is the NYT Books API.

Note: I didn't realize when I did this assignment that it's a deprecated API, so probably wouldn't have chosen that one.

### Load the Libraries

```{r}
library(httr2)
library(purrr)
library(dplyr)
library(tidyr)
library(jsonlite)
library(stringr)
library(tidyjson)
library(tidytext)
library(tidyverse)
```

### Read the Data

Read the current best sellers list for hardcover fiction:

```{r}
# API Call
nyt_url <- "https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json"

# Get the API Key
api <- read_csv("https://raw.githubusercontent.com/gillianmcgovern0/cuny-data-607/refs/heads/main/NYT_API.csv")
api_key <- api$api_key

# Must include API key in the URL
full_url <- paste(nyt_url, api_key, sep = "?api-key=")

# API Request
req <- fromJSON(full_url, flatten=TRUE)

# Actual df exists under requests -> books
df <- req$results$books
glimpse(df)
head(df)
```

As we can see above, `isbns` and `buy_links` are both lists, so we can't see the data after using `fromJSON`. So we need to unnest these variables to show the values:

```{r}
# Unnest the variables that are lists or objects
final_df <- df |> 
  unnest_wider(buy_links) |>
  unnest_wider(isbns)
head(final_df)
```

This data frame now looks exactly like the original JSON (we can see all the values).

Yet, having a list for a data frame value is not a good structure. For example, if we wanted to add another buy link for a book, we need to update the lists in `name` and `url` which is not feasible.

The other list variables are `isbn10` and `isbn13`. An ISBN is the International Standard Book Number that identifies a book. According to online research, it looks like `isbn10` (10 digit ID) and `isbn13` (13 digit ID) have no relation to each other, they are separate ISBNs. So for this particular data frame, it makes more sense to have one `isbn` column that includes ISBN10 and ISBN13 values. We could also create separate data frames to separate this data, but I will hold off on doing that for this assignment.

### Tidy the Data Frame

Let's first separate the buy links. The `name` and `url` columns are connected to each other, so let's `unnest` them at the same time to turn each value into a separate row/observation:

```{r}
# Unnest connected columns at the same time to create accurate observations
final_df_wider <- final_df |>
  unnest(c("name", "url"))
head(final_df_wider)
```

We now have each row represent a unique book and store combination.

Now let's combine the ISBN values into one column, then separate into rows:

```{r}
# Create one ISBN variable so we can have ISBN related observations
final_df_wider2 <- final_df_wider |> 
  mutate(isbn = mapply(c, isbn10, isbn13)) |>
  unnest(isbn) # unnest the list

# Remove the old variables
drop <- c("isbn10","isbn13")
final_df_wider2 <- final_df_wider2[ , !(names(final_df_wider2) %in% drop)]

# Final tidy data frame
head(final_df_wider2)
```

We now have a final tidy date frame where each row represents an ISBN (book identifier) and a store where the book is sold. So if we have a new ISBN best seller, we can just add an additional row and not make an addition to a variable.

### Conclusions

Takeaways from this assignment:

* `fromJSON` does a good job importing JSON
* `unnest_wider` and `unnest` makes it very easy to keep the data frame tidy when separating columns into rows
